``Interestingness'' in Mathematical Theory Exploration

### 2019-06-02 PhDThesis/evaluation_existingcode.tex
Fix metadata
Thesis updates

For a more realistic test we also applied recurrent clustering to a variety of
the most-downloaded packages from \textsc{Hackage} (as of 2015-10-30), including
\texttt{text} (as above), \texttt{pandoc}, \texttt{attoparsec},
\texttt{scientific}, \texttt{yesod-core} and \texttt{blaze-html}. Whilst we
expected functions with a similar purpose to appear together, such as the
various reader and writer functions of \hs{pandoc}, there were always a few
exceptions which became separate for reasons which are still unclear.

@@

It is currently infeasible to extend these clustering applications into full
theory exploration runs, due to the use of constraints like \qcheck{}'s
\hs{Arbitrary} type class (or LeanCheck's , in the case of Speculate) to
generate suitable values for instantiating variables. Whilst we can
automatically define \qspec{} theories and invoke them with \hs{nix-eval}, not
all types have \hs{Arbitrary} instances; those without cannot be given any
variables in our signature, which severely limits the possible combinations
which can be explored. In many cases no variables can be included at all,
leaving just equations involving constants. This has so far prevented us from
measuring the direct impact on \qspec{} performance on real code examples,
either directly by exploring the sub-sets identified through recurrent
clustering or indirectly by comparing the equations generated by a full
brute-force search to our recurrent clusters.

The following less-serious problems were also encountered while applying
\textsc{ML4HS} to \textsc{Hackage} packages:

\begin{itemize}
\item Some packages, such as \texttt{warp} and \texttt{conduits}, get no
  declarations to cluster. This is because they make all of their declarations
  privately, e.g. in ``internal'' modules, then use separate modules to export
  the public declarations. GHC's renaming phase makes all references to such
  exports canonical, by pointing them to the private declarations. This forces
  us to ignore such declarations, as \qspec{} will not be able to access them.

@@

It seems like this recurrent clustering method has promise, although it will
require a more thorough exploration of the parameters to obtain more intuitively
reliable results. These clusters can then be used in several ways to perform
theory exploration; the most na\"{\i}ve way being to explore each cluster as
\qspec{} signature in its own right. \textsc{ML4HS} already provides this
functionality, although the lack of test generators severely limits what can be
discovered.

As alluded to previously, we also have the opportunity to reason by
\emph{analogy}. Similar to the work on ACL2(ml) \cite{heras2013proof}, we could
produce a general ``scheme'' from each equation we find (either through \qspec{}
or by data mining test suites); like Isabelle approaches have shown
\cite{Montano-Rivas.McCasland.Dixon.ea:2012}, these schemes could then be
instantiated to a variety of similar values, in an attempt to find new theorems
which are analogues of existing results from a different context. ``Mutating''
existing theorem statements in such a way would also increase the chance of any
result being considered interesting; since it's likely that the unmodified
statement was deemed interesting, and the new result would not in general follow
as a simple logical consequence.

>>>>

For a more realistic test we also applied recurrent clustering to a variety of
the most-downloaded packages from \hackage{} (as of 2015-10-30), including
\texttt{text} (as above), \texttt{pandoc}, \texttt{attoparsec},
\texttt{scientific}, \texttt{yesod-core} and \texttt{blaze-html}. Whilst we
expected functions with a similar purpose to appear together, such as the
various reader and writer functions of \hs{pandoc}, there were always a few
exceptions which became separate for reasons which are still unclear.

@@

It is currently infeasible to extend these clustering applications into full
theory exploration runs, due to the use of constraints like \quickcheck{}'s
\hs{Arbitrary} type class (or \leancheck{}'s , in the case of \speculate{}) to
generate suitable values for instantiating variables. Whilst we can
automatically define \quickspec{} theories and invoke them with \hs{nix-eval},
not all types have \hs{Arbitrary} instances; those without cannot be given any
variables in our signature, which severely limits the possible combinations
which can be explored. In many cases no variables can be included at all,
leaving just equations involving constants. This has so far prevented us from
measuring the direct impact on \quickspec{} performance on real code examples,
either directly by exploring the sub-sets identified through recurrent
clustering or indirectly by comparing the equations generated by a full
brute-force search to our recurrent clusters.

The following less-serious problems were also encountered while applying
\mlforhs{} to \hackage{} packages:

\begin{itemize}
\item Some packages, such as \texttt{warp} and \texttt{conduits}, get no
  declarations to cluster. This is because they make all of their declarations
  privately, e.g. in ``internal'' modules, then use separate modules to export
  the public declarations. GHC's renaming phase makes all references to such
  exports canonical, by pointing them to the private declarations. This forces
  us to ignore such declarations, as \quickspec{} will not be able to access them.

@@

It seems like this recurrent clustering method has promise, although it will
require a more thorough exploration of the parameters to obtain more intuitively
reliable results. These clusters can then be used in several ways to perform
theory exploration; the most na\"{\i}ve way being to explore each cluster as
\quickspec{} signature in its own right. \mlforhs{} already provides this
functionality, although the lack of test generators severely limits what can be
discovered.

As alluded to previously, we also have the opportunity to reason by
\emph{analogy}. Similar to the work on ACL2(ml) \cite{heras2013proof}, we could
produce a general ``scheme'' from each equation we find (either through \quickspec{}
or by data mining test suites); like Isabelle approaches have shown
\cite{Montano-Rivas.McCasland.Dixon.ea:2012}, these schemes could then be
instantiated to a variety of similar values, in an attempt to find new theorems
which are analogues of existing results from a different context. ``Mutating''
existing theorem statements in such a way would also increase the chance of any
result being considered interesting; since it's likely that the unmodified
statement was deemed interesting, and the new result would not in general follow
as a simple logical consequence.

### 2019-06-02 PhDThesis/futurework.tex
Fix metadata of listings
Thesis updates

Our use of clustering to pre-process \qspec{} signatures has required many
decisions and tradeoffs to be made. Hence our approach is just one possibility
out of many alternatives which could be investigated to push this work
further. In addition, there are other ways in which machine learning could aid
theory exploration besides our relevance filter technique. Below, we elaborate
on the details, background and motivation for these choices.

@@

One partial solution would be leave most Core expressions without types, but to
include them for non-local identifiers (i.e. globals and constructors), which we
can look up in a database. In fact, our ML4HS framework already includes such
type information in its database, alongside the Core syntax trees. Integrating
this information into our algorithm is the next logical step.

@@

Our current approach is a rather conservative change to the existing theory
exploration approaches, as it is essentially a wrapper around \qspec{}. There is
potential for more radical changes to be made, which alter the search process
itself.

@@

\qcheck{} is certainly the most popular property checker for Haskell, which
motivates its use in \qspec{} to instantiate variables to random
values. However, this task of finding type inhabitants has also been solved in
many other ways, which may be worth investigating in place of \qcheck{} (or
perhaps even as part of an ensemble).

The \textsc{SmallCheck} system \cite{runciman2008smallcheck} \emph{enumerates}
values rather than sampling them randomly. Whilst this does not make
\textsc{SmallCheck} objectively ``better'' than \qcheck{}, one major advantage
is that it may use much less memory, as the generated values are built up
incrementally. In contrast, \qcheck{} may generate very large values; in
particular, generating tree structures na\"{\i}vely can cause them to grow
exponentially. For example, here is a potential generator for \hs{RoseTree}s:

\begin{lstlisting}[language=Haskell, xleftmargin=0.1\textwidth, xrightmargin=0.1\textwidth]
genRoseTree = do f        <- arbitrary
                 subtrees <- listOf genRoseTree
                 return (Node f subtrees)
\end{lstlisting}

@@

A relative of \textsc{SmallCheck} is \textsc{Lazy SmallCheck}
\cite{reich2013advances}, which uses laziness to only produce parts of a
datastructure as they are demanded. This may narrow down our search procedures
greatly, especially when predicates are involved. \qcheck{} allows predicates to
restrict the values it tests with, and hence allows \emph{conditional} equations
to be discovered. However, its implementation uses a simple rejection sampling
technique: values are generated just as if the predicate were not there, and
afterwards are filtered to reject any which do not satisfy the predicate. This
makes it difficult to use very specific predicates, as it is unlikely that many
of our random samples will exactly match our criteria. On the other hand,
\textsc{Lazy SmallCheck} will focus its search on exactly those parts of the
datastructure which are checked by the predicate, as those are the parts being
forced to evaluate. This makes it much more likely that we will find values
which satisfy the predicate, allowing us to effectively explore more specific
conditional properties.

Other approaches to generating inhabitants include \textsc{Djinn}
\cite{augustsson2005djinn}, which uses a decision procedure for a sub-set of
Haskell types which in particular can generate and apply functions (unlike the
above tools, which generate values ``bottom-up'' from constructors, and only use
functions when they have been explicitly written in a
generator). \textsc{MuCheck} \cite{le2014mucheck} is designed for
\emph{mutation testing}, and contains combinators for altering functions in
common ways (e.g. changing the order of pattern match clauses); whilst not as
exhaustive as the other approaches, mutating existing values in this way is
claimed to yield values which correspond more closely to what a programmer might
write. This is an interesting possibility for focusing theory exploration on to
more ``realistic'' areas of the search space, and hence avoiding some of the
more useless or bizarre expressions that random search and enumeration may
produce.

In fact, the database generated by our \textsc{AstPlugin} may prove helpful in
generating values, since its type information can be fed to a tool like
\textsc{Djinn} to discover chains of function applications for building values,
which would be particularly useful in cases where constructors are private, like
in our email example. This is similar to the \textsc{Hoogle} tool, but also
offers the ability to use dependency information to avoid potentially infinite
recursion.

The Core syntax trees in our database could also be used to generate theories
for automated theorem provers. \hspec{} currently uses the GHC API to transform
Core within its own process, however that approach suffers from the problems
described in \S \ref{sec:astplugin}.

@@

This is governed by the ``interestingness'' criteria of the theory exploration
system: what to keep and what to discard, and even what areas of the search
space to prioritise. \qspec{}'s approach, briefly mentioned in \S
\ref{sec:theoryexploration}, is very simple: we discard equations which are
direct consequences of others, and keep all the rest. Different, and more
sophisticated notions of interestingness have been widely studied in other
fields, which may be applied in the context of theory exploration.

@@

Approaches vary, from those which are directly related to theory exploration
(such as the scheme-instantiating approach of
\cite{Montano-Rivas.McCasland.Dixon.ea:2012}, which forms part of a theory
exploration system), to others which are more closely related to theories of
human learning and discovery \cite{Piantadosi.Tenenbaum.Goodman:2012,
  mullerunderstanding}. Those based on finding patterns in data, such as
\cite{Wille:2005}, may be useful in tandem with our expression database, and
the results of value generators like \qcheck{}.

Since tools like \hspec{} already call out to third-party automated theorem
provers, and indeed our own ML4HS system uses the external Weka program, there
may also be merit in using external concept formation or conjecture generation
tools (or reimplementations of their ideas), in order to build up more structure
on top of that provided by the code we are exploring. For example, the
approaches taken by AM \cite{lenat1977automated, lenat1979automated}, Graffiti
\cite{delavina2005some, delavina2005graffiti} and HR
\cite{colton1999automatic, colton2000agent} could be used alongside those of
\qspec{}.

@@

Many of these reward signals are based on information theory, with a prominent
example being \emph{compression progress}: given a compressed representation of
our previous observations, the ``progress'' is the space saved if we include the
current observation. Observations which are incompressible or trivially
compressible don't save any space, whilst observations which provide new
insights into the structure of past experience can provide a space saving when
compressed together. This seems particularly relevant for our problem of
identifying interesting theorems: those new theorems (``observations'') which
shorten the proofs of previously discovered theorems may be more general, more
powerful and therefore more \emph{interesting} and hence worth keeping. In fact
this is very similar to \qspec{}'s interestingness criterion.

>>>>

Our use of clustering to pre-process \quickspec{} signatures has required many
decisions and tradeoffs to be made. Hence our approach is just one possibility
out of many alternatives which could be investigated to push this work
further. In addition, there are other ways in which machine learning could aid
theory exploration besides our relevance filter technique. Below, we elaborate
on the details, background and motivation for these choices.

@@

One partial solution would be leave most Core expressions without types, but to
include them for non-local identifiers (i.e. globals and constructors), which we
can look up in a database. In fact, our \mlforhs{} framework already includes such
type information in its database, alongside the Core syntax trees. Integrating
this information into our algorithm is the next logical step.

@@

Our current approach is a rather conservative change to the existing theory
exploration approaches, as it is essentially a wrapper around \quickspec{}.
There is potential for more radical changes to be made, which alter the search
process itself.

@@

\quickcheck{} is certainly the most popular property checker for Haskell, which
motivates its use in \quickspec{} to instantiate variables to random
values. However, this task of finding type inhabitants has also been solved in
many other ways, which may be worth investigating in place of \quickcheck{} (or
perhaps even as part of an ensemble).

The \smallcheck{} system \cite{runciman2008smallcheck} \emph{enumerates}
values rather than sampling them randomly. Whilst this does not make
\smallcheck{} objectively ``better'' than \quickcheck{}, one major advantage
is that it may use much less memory, as the generated values are built up
incrementally. In contrast, \quickcheck{} may generate very large values; in
particular, generating tree structures na\"{\i}vely can cause them to grow
exponentially. For example, here is a potential generator for \hs{RoseTree}s:

\begin{haskell}
genRoseTree = do f        <- arbitrary
                 subtrees <- listOf genRoseTree
                 return (Node f subtrees)
\end{haskell}

@@

A relative of \smallcheck{} is \lazysmallcheck{}~\cite{reich2013advances}, which
uses laziness to only produce parts of a datastructure as they are demanded.
This may narrow down our search procedures greatly, especially when predicates
are involved. \quickcheck{} allows predicates to restrict the values it tests
with, and hence allows \emph{conditional} equations to be discovered. However,
its implementation uses a simple rejection sampling technique: values are
generated just as if the predicate were not there, and afterwards are filtered
to reject any which do not satisfy the predicate. This makes it difficult to use
very specific predicates, as it is unlikely that many of our random samples will
exactly match our criteria. On the other hand, \lazysmallcheck{} will focus its
search on exactly those parts of the datastructure which are checked by the
predicate, as those are the parts being forced to evaluate. This makes it much
more likely that we will find values which satisfy the predicate, allowing us to
effectively explore more specific conditional properties.

Other approaches to generating inhabitants include
\djinn{}~\cite{augustsson2005djinn}, which uses a decision procedure for a sub-set of
Haskell types which in particular can generate and apply functions (unlike the
above tools, which generate values ``bottom-up'' from constructors, and only use
functions when they have been explicitly written in a
generator). \mucheck{}~\cite{le2014mucheck} is designed for
\emph{mutation testing}, and contains combinators for altering functions in
common ways (e.g. changing the order of pattern match clauses); whilst not as
exhaustive as the other approaches, mutating existing values in this way is
claimed to yield values which correspond more closely to what a programmer might
write. This is an interesting possibility for focusing theory exploration on to
more ``realistic'' areas of the search space, and hence avoiding some of the
more useless or bizarre expressions that random search and enumeration may
produce.

In fact, the database generated by our \astplugin{} may prove helpful in
generating values, since its type information can be fed to a tool like
\djinn{} to discover chains of function applications for building values,
which would be particularly useful in cases where constructors are private, like
in our email example. This is similar to the \hoogle{} tool, but also offers the
ability to use dependency information to avoid potentially infinite recursion.

The Core syntax trees in our database could also be used to generate theories
for automated theorem provers. \hipspec{} currently uses the GHC API to transform
Core within its own process, however that approach suffers from the problems
described in \S~\ref{sec:astplugin}.

@@

This is governed by the ``interestingness'' criteria of the theory exploration
system: what to keep and what to discard, and even what areas of the search
space to prioritise. \quickspec{}'s approach, briefly mentioned in
\S~\ref{sec:theoryexploration}, is very simple: we discard equations which are
direct consequences of others, and keep all the rest. Different, and more
sophisticated notions of interestingness have been widely studied in other
fields, which may be applied in the context of theory exploration.

@@

Approaches vary, from those which are directly related to theory exploration
(such as the scheme-instantiating approach of
\cite{Montano-Rivas.McCasland.Dixon.ea:2012}, which forms part of a theory
exploration system), to others which are more closely related to theories of
human learning and discovery \cite{Piantadosi.Tenenbaum.Goodman:2012,
  mullerunderstanding}. Those based on finding patterns in data, such as
\cite{Wille:2005}, may be useful in tandem with our expression database, and
the results of value generators like \quickcheck{}.

Since tools like \hipspec{} already call out to third-party automated theorem
provers, and indeed our own \mlforhs{} system uses the external Weka program, there
may also be merit in using external concept formation or conjecture generation
tools (or reimplementations of their ideas), in order to build up more structure
on top of that provided by the code we are exploring. For example, the
approaches taken by AM \cite{lenat1977automated, lenat1979automated}, Graffiti
\cite{delavina2005some, delavina2005graffiti} and HR
\cite{colton1999automatic, colton2000agent} could be used alongside those of
\quickspec{}.

@@

Many of these reward signals are based on information theory, with a prominent
example being \emph{compression progress}: given a compressed representation of
our previous observations, the ``progress'' is the space saved if we include the
current observation. Observations which are incompressible or trivially
compressible don't save any space, whilst observations which provide new
insights into the structure of past experience can provide a space saving when
compressed together. This seems particularly relevant for our problem of
identifying interesting theorems: those new theorems (``observations'') which
shorten the proofs of previously discovered theorems may be more general, more
powerful and therefore more \emph{interesting} and hence worth keeping. In fact
this is very similar to \quickspec{}'s interestingness criterion.

### 2019-06-02 PhDThesis/litreview_automatedreasoning.tex
Fix references
Headingy things
Thesis updates

\subsection{Automated Mathematical Discovery and Invention}

@@

A similar ``layering'' approach is found in the \textsc{IsaScheme} system of
Monta{\~n}o-Rivas \etal{}~\cite{Montano-Rivas.McCasland.Dixon.ea:2012}, which
has also been quantitatively compared against \textsc{IsaCoSy} and
\textsc{HipSpec} using precision/recall analysis~\cite{claessen2013automating}.
The name comes from its embedding in the Isabelle proof assistant and its use of
``schemes'': higher-order formulae which can be used to generate new concepts
and conjectures. Variables within a scheme are instantiated automatically and
this drives the invention process. For example, the concept of ``repetition''
can be encoded as a scheme, and instantiated with existing encodings of zero,
successor and addition to produce a definition of multiplication. The same
scheme can be instantiated with this new multiplication function to produce
exponentiation.

\textsc{IsaCoSy} (Isabelle Conjecture Synthesis) is written for the Isabelle
proof assistant, mostly in
StandardML~\cite{Johansson.Dixon.Bundy:conjecture-generation}. It conjectures
equations by enumerating expressions involving a given set of (typed) constants
and free variables (a \emph{signature}). A constraint solver forbids certain
(sub)expressions from being synthesised, and these constraints are extended
whenever a new property is conjectured, to avoid generating any special-cases of
this property in the future. Conjectures are tested by looking for
counterexamples and, if none are found, sent to \textsc{IsaPlanner} which
attempts to prove them.

\textsc{QuickSpec} (the conjecture generation component of \textsc{HipSpec})
emerged from work on the \textsc{QuickCheck} software testing framework for the
Haskell programming language~\cite{claessen2011quickcheck}. As well as
conjecturing properties of Haskell definitions for \textsc{HipSpec}, it has been
applied to Isabelle definitions via the \textsc{Hipster} tool~\cite{Hipster}.
Like \textsc{IsaCoSy}, \textsc{QuickSpec} takes a signature and enumerates
well-typed terms. It collects together those of the same type into equivalence
classes, assuming them to be equal, then uses \textsc{QuickCheck} to find
counterexamples to this assumption by randomly instantiating the variables and
comparing the resulting closed expressions. Any equivalence class whose elements
don't compare equal are split up, and the process is repeated with new random
values.

After 500 rounds of testing, any expressions still sharing the same equivalence
class are conjectured to be equal for all values of their variables. Rather than
using a constraint system to prevent redundancies (such as special-cases of
other properties), \textsc{QuickSpec} instead passes its output through a
congruence closure algorithm to achieve the same effect.

\textsc{QuickSpec} has since evolved to version 2~\cite{smallbone2017quick},
which replaces the distinct enumeration and testing steps with a single,
iterative algorithm similar to that of \textsc{IsaCoSy}. Generated conjectures
are fed into a Knuth-Bendix completion algorithm to form a corresponding set of
rewrite rules. As expressions are enumerated, they are simplified using these
rules and discarded if equal to a known expression. If not, \textsc{QuickCheck}
tests whether the new expression can be distinguished from the known expressions
through random testing: those which can are added to the set of known
expressions. Those which cannot be distinguished are conjectured to be equal,
and the rewrite rules are updated.

\textsc{QuickSpec} has also inspired another MTE tool for Haskell called
\textsc{Speculate}~\cite{braquehais2017speculate}, which operates in a similar
way but also makes use of the laws of total orders and Boolean algebra to
conjecture \emph{in}equalities and conditional relations between expressions.

>>>>

\section{Automated Mathematical Discovery and Invention}

@@

A similar ``layering'' approach is found in the \isascheme{} system of
Monta{\~n}o-Rivas \etal{}~\cite{Montano-Rivas.McCasland.Dixon.ea:2012}, which
has also been quantitatively compared against \isacosy{} and \hipspec{} using
precision/recall analysis~\cite{claessen2013automating}.
The name comes from its embedding in the Isabelle proof assistant and its use of
``schemes'': higher-order formulae which can be used to generate new concepts
and conjectures. Variables within a scheme are instantiated automatically and
this drives the invention process. For example, the concept of ``repetition''
can be encoded as a scheme, and instantiated with existing encodings of zero,
successor and addition to produce a definition of multiplication. The same
scheme can be instantiated with this new multiplication function to produce
exponentiation.

\isacosy{} (Isabelle Conjecture Synthesis) is written for the Isabelle
proof assistant, mostly in
StandardML~\cite{Johansson.Dixon.Bundy:conjecture-generation}. It conjectures
equations by enumerating expressions involving a given set of (typed) constants
and free variables (a \emph{signature}). A constraint solver forbids certain
(sub)expressions from being synthesised, and these constraints are extended
whenever a new property is conjectured, to avoid generating any special-cases of
this property in the future. Conjectures are tested by looking for
counterexamples and, if none are found, sent to \textsc{IsaPlanner} which
attempts to prove them.

\quickspec{} (the conjecture generation component of \hipspec{})
emerged from work on the \quickcheck{} software testing framework for the
Haskell programming language~\cite{claessen2011quickcheck}. As well as
conjecturing properties of Haskell definitions for \hipspec{}, it has been
applied to Isabelle definitions via the \hipster{} tool~\cite{Hipster}.
Like \isacosy{}, \quickspec{} takes a signature and enumerates
well-typed terms. It collects together those of the same type into equivalence
classes, assuming them to be equal, then uses \quickcheck{} to find
counterexamples to this assumption by randomly instantiating the variables and
comparing the resulting closed expressions. Any equivalence class whose elements
don't compare equal are split up, and the process is repeated with new random
values.

After 500 rounds of testing, any expressions still sharing the same equivalence
class are conjectured to be equal for all values of their variables. Rather than
using a constraint system to prevent redundancies (such as special-cases of
other properties), \quickspec{} instead passes its output through a
congruence closure algorithm to achieve the same effect.

\quickspec{} has since evolved to version 2~\cite{smallbone2017quick},
which replaces the distinct enumeration and testing steps with a single,
iterative algorithm similar to that of \isacosy{}. Generated conjectures
are fed into a Knuth-Bendix completion algorithm to form a corresponding set of
rewrite rules. As expressions are enumerated, they are simplified using these
rules and discarded if equal to a known expression. If not, \quickcheck{}
tests whether the new expression can be distinguished from the known expressions
through random testing: those which can are added to the set of known
expressions. Those which cannot be distinguished are conjectured to be equal,
and the rewrite rules are updated.

\quickspec{} has also inspired another MTE tool for Haskell called
\speculate{}~\cite{braquehais2017speculate}, which operates in a similar
way but also makes use of the laws of total orders and Boolean algebra to
conjecture \emph{in}equalities and conditional relations between expressions.

### 2019-06-02 PhDThesis/litreview_interestingness.tex
Thesis updates

% FIXME: This is original work, so should be in a different section
\begin{table}
  \centering
  \begin{tabular}{ |l|l|c|c|c|c|c|c| }
    \hline
    \multirow{2}{*}{\textbf{Program}}                      &
    \multirow{2}{*}{\textbf{Conjecture Types}}             &
    \multicolumn{6}{c|}{\textbf{Interestingness Measures}} \\ \hhline{~~------}
    \tRow{          &                     & \iE & \iN & \iS & \iA & \iC & \iU}
    \tRow{AM        & \tIFF, \tIMP, \tNE  &   X &   X &   X &   X &   X &   X}
    \tRow{GT        & \tIFF, \tIMP, \tNE  &   X &   X &   X &   X &   X &   X}
    \tRow{Graffiti  & \tINE               &   X &   X &   X &     &   X &   X}
    \tRow{\Bagai{}  & \tNE                &     &   X &     &   X &   X &    }
    \tRow{HR        & \tIFF, \tIMP, \tNE  &   X &   X &   X &   X &   X &   X}
    \tRow{QuickSpec & \tEQ                &   X &   X &     &   X &   X &    }
    \tRow{Speculate & \tCON\ \tEQ / \tINE &   X &   X &     &   X &   X &    }
    \tRow{IsaCoSy   & \tEQ                &   X &   X &     &   X &   X &   X}
    \tRow{IsaScheme & \tEQ                &   X &   X &     &   X &   X &   X}
  \end{tabular}
  \caption{Classification of MTE tools from~\cite{colton2000notion}, extended
    to include four more recent tools. The interestingness measures are
    \iE{}mpirical plausibility, \iN{}ovelty, \iS{}urprisingness,
    \iA{}pplicability, \iC{}omprehensibility (low complexity) and \iU{}tility.}
  \label{table:colton}
\end{table}

% JUSTIFICATIONS
%
% IsaCoSy uses counter-example checking; this ensures empirical plausibility
% IsaCoSy uses constraints to avoid special-cases; this ensures novelty
% IsaCoSy uses utility, since there are some hard-coded patterns which are
% looked for
%
% Speculate uses a form of unification to ensure novelty, making sure one
% equation is not a special-case of another
% Speculate uses LeanCheck to enumerate values, looking for counterexamples
% Speculate uses testing, to ensure Empirical Plausibility
%
% QuickSpec uses QuickCheck to ensure empirical plausibility
% QuickSpec uses a congruence closure algorithm to ensure novelty
%
% IsaScheme uses utility, since it determines the "quality" of a definition
% based on how many theorems it appears in?
% IsaScheme uses novelty, using an equational rewrite system (Knuth-Bendix
% completion) to remove redundancies
% IsaScheme uses utility, since it focuses on the terms and definitions of a
% user's theory. However, don't all of them?
% IsaScheme uses counterexample checking for empirical plausibility.

@@

We summarise the criteria in Table~\ref{table:colton} and state which were used
in key ATF/MTE tools. The entries for AM, GT, Graffiti, \Bagai{} and HR are
based on this survey (the latter is Colton's own tool). We extend this analysis
to some more recent tools in the remaining rows (based on our understanding of
their function): \qspec{}~\cite{QuickSpec},
Speculate~\cite{braquehais2017speculate},
IsaCoSy~\cite{Johansson.Dixon.Bundy:conjecture-generation}
and IsaScheme~\cite{MontanoRivas2011}.

@@

IsaCoSy ensures novelty using a constraint system to prevent special-cases being
generated, whilst the others use post-hoc filters based on unification and term
rewriting to achieve the same result. IsaCoSy is able to look for commonly
desirable patterns, such as commutativity and associativity, which increases the
utility of its output, whilst IsaScheme is completely based around such
pattern-instantiation.

This diversity of approaches makes it difficult to compare the existing
evaluations of these tools directly. In particular, evaluation methods created
for one tool might not make sense for another, due to assumptions made about
the algorithms' operation. For example, the novelty filters used by \qspec{}
and Speculate are applied to the whole output set, guaranteeing that no
special-cases will appear; yet we cannot assume this in the case of IsaCoSy,
since its constraint solver allows special cases if they're found \emph{before}
the general case (it does not go back and discard previous results).

>>>>

% FIXME: This is original work, so should be in a different section
\begin{table}
  \centering
  \begin{tabular}{ |l|l|c|c|c|c|c|c| }
    \hline
    \multirow{2}{*}{\textbf{Program}}                      &
    \multirow{2}{*}{\textbf{Conjecture Types}}             &
    \multicolumn{6}{c|}{\textbf{Interestingness Measures}} \\ \hhline{~~------}
    \tRow{             &                     & \iE & \iN & \iS & \iA & \iC & \iU}
    \tRow{AM           & \tIFF, \tIMP, \tNE  &   X &   X &   X &   X &   X &   X}
    \tRow{GT           & \tIFF, \tIMP, \tNE  &   X &   X &   X &   X &   X &   X}
    \tRow{Graffiti     & \tINE               &   X &   X &   X &     &   X &   X}
    \tRow{\Bagai{}     & \tNE                &     &   X &     &   X &   X &    }
    \tRow{HR           & \tIFF, \tIMP, \tNE  &   X &   X &   X &   X &   X &   X}
    \tRow{\quickspec{} & \tEQ                &   X &   X &     &   X &   X &    }
    \tRow{\speculate{} & \tCON\ \tEQ / \tINE &   X &   X &     &   X &   X &    }
    \tRow{\isacosy{}   & \tEQ                &   X &   X &     &   X &   X &   X}
    \tRow{\isascheme{} & \tEQ                &   X &   X &     &   X &   X &   X}
  \end{tabular}
  \caption{Classification of MTE tools from~\cite{colton2000notion}, extended
    to include four more recent tools. The interestingness measures are
    \iE{}mpirical plausibility, \iN{}ovelty, \iS{}urprisingness,
    \iA{}pplicability, \iC{}omprehensibility (low complexity) and \iU{}tility.}
  \label{table:colton}
\end{table}

% JUSTIFICATIONS
%
% ICoSy uses counter-example checking; this ensures empirical plausibility
% ICoSy uses constraints to avoid special-cases; this ensures novelty
% ICoSy uses utility, since there are some hard-coded patterns which are
% looked for
%
% Speculat uses a form of unification to ensure novelty, making sure one
% equation is not a special-case of another
% Speculat uses LeanCheck to enumerate values, looking for counterexamples
% Speculat uses testing, to ensure Empirical Plausibility
%
% QSpec uses QCheck to ensure empirical plausibility
% QSpec uses a congruence closure algorithm to ensure novelty
%
% IScheme uses utility, since it determines the "quality" of a definition
% based on how many theorems it appears in?
% IScheme uses novelty, using an equational rewrite system (Knuth-Bendix
% completion) to remove redundancies
% IScheme uses utility, since it focuses on the terms and definitions of a
% user's theory. However, don't all of them?
% IScheme uses counterexample checking for empirical plausibility.

@@

We summarise the criteria in Table~\ref{table:colton} and state which were used
in key ATF/MTE tools. The entries for AM, GT, Graffiti, \Bagai{} and HR are
based on this survey (the latter is Colton's own tool). We extend this analysis
to some more recent tools in the remaining rows (based on our understanding of
their function): \quickspec{}~\cite{QuickSpec},
\speculate{}~\cite{braquehais2017speculate},
\isacosy{}~\cite{Johansson.Dixon.Bundy:conjecture-generation}
and \isascheme{}~\cite{MontanoRivas2011}.

@@

\isacosy{} ensures novelty using a constraint system to prevent special-cases
being generated, whilst the others use post-hoc filters based on unification and
term rewriting to achieve the same result. \isacosy{} is able to look for
commonly desirable patterns, such as commutativity and associativity, which
increases the utility of its output, whilst \isascheme{} is completely based
around such pattern-instantiation.

This diversity of approaches makes it difficult to compare the existing
evaluations of these tools directly. In particular, evaluation methods created
for one tool might not make sense for another, due to assumptions made about
the algorithms' operation. For example, the novelty filters used by \quickspec{}
and \speculate{} are applied to the whole output set, guaranteeing that no
special-cases will appear; yet we cannot assume this in the case of \isacosy{},
since its constraint solver allows special cases if they're found \emph{before}
the general case (it does not go back and discard previous results).

### 2019-06-02 PhDThesis/litreview_staticanalysis.tex
Headingy things
Thesis updates

\subsection{Automated Software Improvement}

@@

We can reduce the amount of unknown parameters in a concrete program by
propagating statically-known values, for example using
\emph{supercompilation}~\cite{Turchin:1986:CS:5956.5957}; however this can
dramatically increase code size, as each function may be duplicated into
multiple specialised instances. Related techniques, like \emph{symbolic
  execution} (a form of \emph{abstract interpretation}), allow approximate
execution of code, where unknown runtime data is represented as opaque
metavariables. Property- and model-checkers are similar, except they work
empirically by performing multiple runs, instantiating metavariables with
synthetic data (\qcheck{} does this randomly, \textsc{SmallCheck} enumerates
values, etc.).

@@

The open-ended search process of such superoptimisers is similar to application
of mathematical theory exploration to software property discovery, especially
when exploration is limited to equality properties (which is the case for
\textsc{IsaCoSy} and \textsc{QuickSpec}). Quality metrics like efficiency (e.g.
relative instruction count) can be considered a subset of the interestingness
criteria of theory exploration. The \emph{aims} of these analyses are in
opposition, since theory exploration \emph{builds up} structure and abstraction
latent in a set of definitions, whilst superoptimisation \emph{tears down} such
layers to find minimal implementations; yet the practical necessity of
superoptimising only small snippets at a time (e.g. to find ``peephole
optimisations''~\cite{Bansal.Aiken:2006}) results in remarkably similar
discovery processes.

>>>>

\section{Automated Software Improvement}

@@

We can reduce the amount of unknown parameters in a concrete program by
propagating statically-known values, for example using
\emph{supercompilation}~\cite{Turchin:1986:CS:5956.5957}; however this can
dramatically increase code size, as each function may be duplicated into
multiple specialised instances. Related techniques, like \emph{symbolic
  execution} (a form of \emph{abstract interpretation}), allow approximate
execution of code, where unknown runtime data is represented as opaque
metavariables. Property- and model-checkers are similar, except they work
empirically by performing multiple runs, instantiating metavariables with
synthetic data (\quickcheck{} does this randomly, \smallcheck{} enumerates
values, etc.).

@@

The open-ended search process of such superoptimisers is similar to application
of mathematical theory exploration to software property discovery, especially
when exploration is limited to equality properties (which is the case for
\isacosy{} and \quickspec{}). Quality metrics like efficiency (e.g.
relative instruction count) can be considered a subset of the interestingness
criteria of theory exploration. The \emph{aims} of these analyses are in
opposition, since theory exploration \emph{builds up} structure and abstraction
latent in a set of definitions, whilst superoptimisation \emph{tears down} such
layers to find minimal implementations; yet the practical necessity of
superoptimising only small snippets at a time (e.g. to find ``peephole
optimisations''~\cite{Bansal.Aiken:2006}) results in remarkably similar
discovery processes.

% TODO: Evolutionary patches
% TODO: Hutter search
% TODO: Goedel machines
